---
title: "Lab-4"
author: "YangXu"
format:
  html:
    embed-resources: true
---

1. Read in the data
```{r}
library(data.table)
library(leaflet)
library(tidyverse)
if (!file.exists("met_all.gz"))
  download.file(
    url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz",
    destfile = "met_all.gz",
    method   = "libcurl",
    timeout  = 60
    )
met <- data.table::fread("met_all.gz")
```


2. Prepare the data
```{r}
dim(met)
# Remove temp less than -17C
met <- met[met$temp > -17]

# check if any missing data coded as 9999,999,etc
summary(met)

# Replace missing data in elev with NA
met <- met[elev == 9999, elev:=NA]

# Generate a date variable
met[, week := week(as.Date(paste(year, month, day, sep = "-")))]
met <- met[week == min(week, na.rm = TRUE)]
# Compute the mean
met_avg <- met[,.(
  temp     = mean(temp,na.rm = TRUE),
  rh       = mean(rh,na.rm = TRUE),
  wind.sp  = mean(wind.sp,na.rm=TRUE),
  vis.dist = mean(vis.dist,na.rm=TRUE),
  dew.point = mean(dew.point, na.rm=TRUE),
  lat      = mean(lat),
  lon      = mean(lon), 
  elev     = mean(elev,na.rm=TRUE)
),by = c("USAFID","day")]

# Create region variables
met_avg$region <- ifelse(met_avg$lon > -98 & met_avg$lat >39.71, "NE",
                  ifelse(met_avg$lon > -98 & met_avg$lat < 39.71, "SE",
                  ifelse(met_avg$lon < -98 & met_avg$lat >39.71, "NW", 
                         "SW")))
# Create a categorical variable for elevation
met_avg[, elev_cat := fifelse(elev > 252, "high", "low")]
```

3. Examine the wind speed and dew point by region
```{r}
# Use geom_violin
met_avg[!is.na(region)] %>%
ggplot()+
  geom_violin(mapping = aes(y=wind.sp, x=1)) +
  facet_wrap(~region, nrow=2)
# Use geom_boxplot
met_avg[!is.na(region)] %>%
ggplot()+
  geom_boxplot(mapping = aes(y=wind.sp, fill=region)) +
  facet_wrap(~region, nrow=2)
```
A violin plot is a mirrored density plot displayed in the same way as a boxplot.
From the violin plot, we observe that NE and SE have lower average wind speed than
SW and NW. The distribution of wind speed in SE and NE are less dispersed than SW 
and NW.

4. Examine the association between dew point and wind speed by region
```{r}
met_avg[!is.na(region)] %>%
  ggplot(mapping = aes(x=dew.point, y=wind.sp, color=region))+
  geom_jitter() + 
  stat_smooth(method=lm, se = FALSE)
```
From the plot, we observe that in NW, NE and SE, with dew point increasing, the average 
wind speed gets higher. But in SW, with dew point increasing, the average wind speed
gets lower.


5. Create barplots of the weather stations by elevation category colored by region
```{r}
met_avg[!is.na(region)] %>%
  ggplot()+
  geom_bar(mapping=aes(x=elev_cat,fill=region), position = "dodge")+
  scale_fill_brewer(palette = "BrBG")+
  labs(title="Number of weather stations by elevation category colored by region")+
  labs(x="Elevation Category", y= "Weather Stations")
```
From the plot, we observe that in high elevation area, NE has the most weather stations.
In low elevation area, SE has the most weather stations.


6. Examine mean dew point and wind speed by region with standard deviation error bars
```{r}
met_avg[!is.na(region)] %>%
  ggplot(mapping=aes(x=region, y=dew.point)) +
  stat_summary(fun.data="mean_sdl", geom="errorbar") +
  stat_summary(fun.data="mean_sdl")
met_avg[!is.na(region)] %>%
  ggplot(mapping=aes(x=region, y=wind.sp)) +
  stat_summary(fun.data="mean_sdl", geom="errorbar") +
  stat_summary(fun.data="mean_sdl")
```
SE has the highest mean dew point. SW has the highest mean wind speed.


7. Make a map showing the spatial trend in relative humidity in the US
```{r}
met_avg2 <- met_avg[!is.na(rh)]
# color plaette
rh.pal <- colorNumeric(c('darkgreen','goldenrod','brown'), domain=met_avg2$rh)
# Top10 places
top10 <- met_avg2[rank(-rh) <= 10]
leaflet(met_avg2) %>%
  addProviderTiles('OpenStreetMap') %>%
  addCircles(lat=~lat, lng=~lon, color=~rh.pal(rh), label=~paste0(round(rh,2), ' rh'), opacity=1,fillOpacity=1, radius=500) %>%
  addMarkers(lat=~lat, lng=~lon, label=~paste0(round(rh,2), ' rh'), data = top10) %>%
  addLegend('bottomleft',pal=rh.pal, values=met_avg2$rh, title="Relative Humidity", opacity=1)
```
Mid-west area of the US has the average lowest RH across the country. The closer to the east and west coasts, the higher the RH.


8. Use a ggplot extension
```{r}
library(patchwork)
p1 <- met_avg[!is.na(region)] %>%
ggplot()+
  geom_violin(mapping = aes(y=wind.sp, x=1)) +
  facet_wrap(~region, nrow=2)

p2 <- met_avg[!is.na(region)] %>%
  ggplot(mapping=aes(x=region, y=dew.point)) +
  stat_summary(fun.data="mean_sdl", geom="errorbar") +
  stat_summary(fun.data="mean_sdl")
p1 + p2
```